

---
title: "Assignment 1 – Guil David"
author: "David Guil (david.guil01@estudiant.upf.edu)"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
    theme: flatly
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This practical focuses on data wrangling, exploratory data analysis, visualization, and critical comparison with published results using real biomedical data.

We will work with the publication [**Proteomic and Metabolomic Characterization of COVID-19 Patient Sera**](https://www.sciencedirect.com/science/article/pii/S0092867420306279)

# Exercise 1  

Download and load Supplementary Table 1 (“Additional Demographical and Baseline Characteristics of COVID-19 Patients and Control Groups”) from the supplementary materials of the publication.
 
## Exercise 1.1  

Reproduce Table 1 from the manuscript using the supplementary data. Match the reported summary statistics   as closely as possible.

Briefly discuss any discrepancies between your results and the published table, and provide possible explanations.

```{r}
library(ggplot2)
library(openxlsx)
library(tidyverse)
library(table1)
library(dplyr)
library(rvest)
library(htmltools)


table_1 <- read.xlsx("./0_dataset_raw/table_s1.xlsx", sheet = "Clinical_information")

#Cleaning the database
names(table_1) <- str_to_lower(names(table_1))
table_1[table_1 == "/"] <- NA
table_1$proteomics <- as.factor(table_1$proteomics)
table_1$group.d <- as.factor(table_1$group.d)
table_1$sex.g <- as.factor(table_1$sex.g)
table_1$`covid-19.patients.corhort` <- as.factor(table_1$`covid-19.patients.corhort`)
table_1$`crp.i,.mg/l`<- as.numeric(table_1$`crp.i,.mg/l`)
table_1$bmi.h<- as.numeric(table_1$bmi.h)
table_1$discharge.date <-  as.Date(table_1$discharge.date, origin = "1899-12-30")
table_1$sampling.date.for.metabolomics <- as.Date(table_1$sampling.date.for.metabolomics, origin = "1899-12-30")
table_1$sampling.date.for.proteomics <-  as.Date(as.numeric(table_1$sampling.date.for.proteomics), origin= "1899-12-30")
table_1$test.date <-  as.Date(as.numeric(table_1$test.date), origin= "1899-12-30")
table_1$admission.date <-  as.Date(table_1$admission.date, origin= "1899-12-30")
table_1$onset.date.f <- as.Date(table_1$onset.date.f, origin = "1899-12-30")
table_1$date.of.progression.to.severe.state <-  as.Date(as.numeric(table_1$date.of.progression.to.severe.state), origin= "1899-12-30")

table_1$group.d <- factor(
  table_1$group.d,
  levels = c("0", "1", "2", "3"),
  labels = c("Healthy Control", "Non-COVID-19", "Non-Severe", "Severe"))

table_1$sex.g <- factor(
  table_1$sex.g,
  levels = c("1", "0"),
  labels = c("Male", "Female")
)

table_1$proteomics <- factor(
  table_1$proteomics,
  levels = c("No", "Yes"),
  labels = c("no proteomics", "proteomics available")
)



names(table_1) <- make.unique(names(table_1))

table_1 <- table_1 %>% mutate(
  time_from_onset_to_admission_days = admission.date - onset.date.f
)

table_1 <- table_1 %>% mutate(
  time_from_admission_to_severe_days = date.of.progression.to.severe.state - admission.date
)

table_1$time_from_onset_to_admission_days <- as.numeric(
  table_1$time_from_onset_to_admission_days)

table_1$time_from_admission_to_severe_days <- as.numeric(
  table_1$time_from_admission_to_severe_days
)



#Making a Total COVID group

table_1_total <- table_1 %>%
  filter(`group.d` %in% c("Non-Severe", "Severe")) %>%
  mutate(`group.d` = "Total")

table_1_expanded <- bind_rows(table_1, table_1_total)

table_1_expanded$group.d <- factor(
  table_1_expanded$group.d,
  levels = c(
    "Healthy Control",
    "Non-COVID-19",
    "Total",
    "Non-Severe",
    "Severe"
  )
)


#Labeling the variables for the table
label(table_1_expanded$sex.g) <- "Sex-no.(%)"
label(table_1_expanded$`age.(year)`) <- "Age- year"
label(table_1_expanded$bmi.h) <- "BMI, kg/m²"
label(table_1_expanded$time_from_onset_to_admission_days) <- "Time from Onset to Admission, Days"
label(table_1_expanded$time_from_admission_to_severe_days) <- "Time from Admission to Severe, Days"

#Table Creation
my.render.cont <- function(x) {
  with(stats.apply.rounding(stats.default(x)), {
    out <- c("",
             "Mean ± SD" = sprintf("%s ± %s", MEAN, SD),
             "Median (IQR)" = sprintf("%s (%s–%s)", MEDIAN, Q1, Q3),
             "Range" = sprintf("%s–%s", MIN, MAX))
    out[names(out) != "Missing"]
  })
}

my.render.cat <- function(x) {
  c("", sapply(levels(x), function(lv) {
    n <- sum(x == lv, na.rm = TRUE)
    N <- sum(!is.na(x))
    sprintf("%d (%.1f)", n, 100 * n / N)
  }, USE.NAMES = TRUE))
}


table_output <- table1(  ~ `sex.g` + `age.(year)` + `bmi.h` + `time_from_onset_to_admission_days` + `time_from_admission_to_severe_days`
         | `group.d`,
         data = table_1_expanded,  render.continuous = my.render.cont, render.categorical = my.render.cat, topclass="Rtable1-zebra", overall = FALSE, 
         caption = 'Table 1. Demographics and Baseline Characteristics of COVID-19 Patients'
)

table_df <- as.character(table_output) %>%
  read_html() %>%
  html_table() %>%
  .[[1]]


table_df <- table_df[-c(12,17,22),]

table_df[table_df == "NA"] <- ""
table_df[table_df == "NA ± NA"] <- ""
table_df[table_df == "NA (NA–NA)"] <- ""
table_df[table_df == "NA–NA"] <- ""

html_table1_format <- tags$table(
  class = "Rtable1 Rtable1-zebra",
  tags$thead(
    tags$tr(
      lapply(names(table_df), function(x) tags$th(x))
    )
  ),
  tags$tbody(
    lapply(1:nrow(table_df), function(i) {
      tags$tr(
        lapply(table_df[i, ], function(x) tags$td(x))
      )
    })
  )
)




```

## Exercise 1.2
Perform an exploratory data analysis (EDA) of the dataset. At a minimum, include:

- Descriptive statistics for key variables  
- Appropriate visualizations (e.g. distributions, group comparisons)  
- A brief written summary of notable patterns, anomalies, or data quality issues  


```{r}
library(psych)
library(knitr)
library(kableExtra)
library(dplyr)
library(ggplot2)
library(naniar)

#descriptive statistics
numeric_variables <- table_1 %>% select(where(is.numeric))

numeric_descriptive_table <- psych::describe(numeric_variables)

kable(numeric_descriptive_table, digits = 2, caption = "Descriptive Statistics of Numeric Variables") %>%
  kable_styling(full_width = FALSE, position = "center")

categorical_variables <- table_1 %>% select(where(is.factor))
categorical_descriptive_table <- lapply(categorical_variables, table)

summary(categorical_descriptive_table)
#graph 1
table_1 %>% ggplot(aes(x= group.d , y= `age.(year)`, fill = group.d)) + geom_violin(alpha = 0.7, trim = FALSE) + geom_boxplot(width = 0.1) +
  scale_fill_brewer(palette = "Set2") + theme_minimal() + labs(title = "Age distribution across the diseases group", 
                                                               x = "Diseases group",
                                                               y = "Age") + theme(legend.position = "none", 
                                                               plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

#graph 2
table_1 %>% ggplot(aes(x= group.d , y= bmi.h, fill = group.d)) + geom_violin(alpha = 0.7, trim = FALSE) + geom_boxplot(width = 0.1) +
  scale_fill_brewer(palette = "Set2") + theme_minimal() + labs(title = "BMI distribution across the diseases group", 
                                                               x = "Diseases group",
                                                               y = "BMI") + theme(legend.position = "none", 
                                                               plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

#graph 3
table_1 %>% ggplot(aes(x = group.d, fill = sex.g)) + geom_bar(position = "dodge", alpha = 0.9) + scale_fill_brewer(palette = "Set2", name = "sex") +
  theme_minimal() + labs(title = "Distribution by Disease and Gender", x = "Diseases group", y = "Count", fill= "sex" ) + 
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

#graph 4
miss_table_1 <- vis_miss(table_1)
miss_table_1 + theme(
  axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1, size = 10),
  axis.text.y = element_text(size = 10),
  axis.title = element_text(size = 12),
  plot.margin = margin(5, 5, 5, 5)  # top, right, bottom, left
)

```

# Exercise 2
## Exercise 2.1
Reproduce Supplementary Figure 1 using the data without modifying or removing any observations. Use the same variables and groupings as in the original figure. 
```{r}
library(Hmisc)

table_clinical <- table_1 %>%  select(group_disease = `group.d`,
                                      white_blood_cell = `wbc.count,.×109/l`,
                                      lympocyte = `lymphocyte.count,.×109/l`,
                                      monocyte = `monocyte.count,.×109/l`,
                                      platelet = `platelet.count,.×109/l`,
                                      c_reactive_protein = `crp.i,.mg/l`,
                                      alanine_aminotransferase = `alt.j,.u/l`,
                                      aspartate_aminotransferase = `ast.k,.u/l`,
                                      glutamyltransferase = `ggt.l,.u/l`,
                                      total_bilirubin = `tbil.m,.μmol/l`,
                                      direct_bilirubin = `dbil.n,.μmol/l`,
                                      creatinine = `creatinine,.μmol/l`,
                                      glucose = `glucose,.mmol/l`) %>% filter(group_disease %in% c('Non-COVID-19','Non-Severe', 'Severe'))


table_long <- table_clinical %>%
  pivot_longer(
    cols = -group_disease, #- se usa para elegir todas las columnas, excepto el que va después del signo
    names_to = "parameter",
    values_to = "value"
  )


table_long %>% 
  ggplot(aes(x = group_disease, y = value, fill = group_disease)) +
  stat_summary(fun = mean, geom = "bar", position = "dodge", aes(color = group_disease), alpha = 0.5, width = 0.5) + 
  geom_point(position = position_jitter(width = 0.2), aes(color = group_disease), alpha = 0.5, size = 1) + 
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), 
               geom = "errorbar", width = 0.2, position = position_dodge(0.9)) + 
  facet_wrap(~ parameter, scales = "free_y", ncol = 3) + 
  scale_fill_manual(values = c("white", "white", "white")) +  
  scale_color_manual(values = c("blue", "yellow", "red")) + theme_bw()

```


## Exercise 2.2
Identify and handle outliers using an appropriate and well-justified method (e.g. removal, transformation, or robust statistics).

Reproduce Supplementary Figure 1 after outlier handling and discuss how and why the results change.
```{r}
library(outliers)
# After seeing the data, we conclude that we are going to impute the third quartil to the outliers. The reason is that we don't have a larger n (we cannot remove a lot of values) and some data is realted with several COVID states.

#Defining the function and imputing
impute_outliers_Q3 <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR_val <- Q3 - Q1
  upper <- Q3 + 1.5 * IQR_val
  
  x[x > upper] <- Q3
  return(x)
}
vars_clinicas <- c(
  "white_blood_cell",
  "lympocyte",
  "monocyte",
  "platelet",
  "c_reactive_protein",
  "alanine_aminotransferase",
  "aspartate_aminotransferase",
  "glutamyltransferase",
  "total_bilirubin",
  "direct_bilirubin",
  "creatinine",
  "glucose"
)

table_clinical_imputed <- table_clinical

for (v in vars_clinicas) {
  table_clinical_imputed[[v]] <- impute_outliers_Q3(
    table_clinical[[v]]
  )
}

table_long_imputed <- table_clinical_imputed %>%
  pivot_longer(
    cols = -group_disease, #- se usa para elegir todas las columnas, excepto el que va después del signo
    names_to = "parameter",
    values_to = "value"
  )


table_long_imputed %>% 
  ggplot(aes(x = group_disease, y = value, fill = group_disease)) +
  stat_summary(fun = mean, geom = "bar", position = "dodge", aes(color = group_disease), alpha = 0.5, width = 0.5) + 
  geom_point(position = position_jitter(width = 0.2), aes(color = group_disease), alpha = 0.5, size = 1) + 
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), 
               geom = "errorbar", width = 0.2, position = position_dodge(0.9)) + 
  facet_wrap(~ parameter, scales = "free_y", ncol = 3) + 
  scale_fill_manual(values = c("white", "white", "white")) +  
  scale_color_manual(values = c("blue", "yellow", "red")) + theme_bw()



# Several clinical variables showed right-skewed distributions with extremely high values, especially for inflammatory and liver biomarkers. 
# Since these extreme values may reflect real severe cases rather than measurement errors, observations were not removed.

# Outliers were identified using the interquartile range (IQR) method, defining extreme values as those above Q3 +1.5 (IQR).
# Instead of removing them, outliers were imputed by replacing their values with the third quartile (Q3). 
# This conservative approach limits the influence of extreme values while preserving sample size and overall data structure.
# After outlier handling, Supplementary Figure 1 shows reduced variability and less influence of extreme values on the scale of the plot. 
# The median and overall trends across disease groups remain similar, but group comparisons become clearer. 
# This indicates that extreme values mainly affected dispersion rather than the central tendency of the data.



```

# Exercise 3
Create a heatmap of the biomarker data. Include group and gender as annotation variables.

```{r}

table_clinical_heatmap <- table_1 %>%  select(group_disease = `group.d`,
                                      gender = `sex.g` ,
                                      white_blood_cell = `wbc.count,.×109/l`,
                                      lympocyte = `lymphocyte.count,.×109/l`,
                                      monocyte = `monocyte.count,.×109/l`,
                                      platelet = `platelet.count,.×109/l`,
                                      c_reactive_protein = `crp.i,.mg/l`,
                                      alanine_aminotransferase = `alt.j,.u/l`,
                                      aspartate_aminotransferase = `ast.k,.u/l`,
                                      glutamyltransferase = `ggt.l,.u/l`,
                                      total_bilirubin = `tbil.m,.μmol/l`,
                                      direct_bilirubin = `dbil.n,.μmol/l`,
                                      creatinine = `creatinine,.μmol/l`,
                                      glucose = `glucose,.mmol/l`) %>% filter(group_disease %in% c("Non-COVID-19","Non-Severe", "Severe"))






table_clinical_heatmap_imputed <- table_clinical_heatmap

for (v in vars_clinicas) {
  table_clinical_heatmap_imputed[[v]] <- impute_outliers_Q3(
    table_clinical_heatmap[[v]]
  )
}


table_clinical_heatmap_imputed$gender <- as.factor(table_clinical_heatmap_imputed$gender)
table_clinical_heatmap_imputed$group_disease <- factor(
  table_clinical_heatmap_imputed$group_disease,
  levels = c("Non-COVID-19","Non-Severe", "Severe")
)

 

#Selecting the variables
biomarkers <- c(
  "white_blood_cell",
  "lympocyte",
  "monocyte",
  "platelet",
  "c_reactive_protein",
  "alanine_aminotransferase",
  "aspartate_aminotransferase",
  "glutamyltransferase",
  "total_bilirubin",
  "direct_bilirubin",
  "creatinine",
  "glucose"
)
#Filtering the NA values
for (v in biomarkers) {
  table_clinical_heatmap_imputed[[v]][
    is.na(table_clinical_heatmap_imputed[[v]])
  ] <- median(table_clinical_heatmap_imputed[[v]], na.rm = TRUE)
}


library(ggplot2)
library(tidyr)
library(dplyr)

# Data preparation
heatmap_data_long <- table_clinical_heatmap_imputed %>%
  mutate(sample_id = row_number()) %>%
  pivot_longer(
    cols = all_of(biomarkers),
    names_to = "biomarker",
    values_to = "value"
  ) %>%
  group_by(biomarker) %>%
  mutate(z_score = scale(value)[,1]) %>%
  ungroup()

# Heatmap 
ggplot(heatmap_data_long, aes(x = biomarker, y = factor(sample_id), fill = z_score)) +
  geom_tile(color = "white", linewidth = 0.1) +
  scale_fill_gradient2(
    low = "navy", 
    mid = "white", 
    high = "firebrick3", 
    midpoint = 0, 
    name = "Z-score"
  ) +
  facet_grid(group_disease + gender ~ ., scales = "free_y", space = "free_y") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 10),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    strip.text.y = element_text(angle = 0, hjust = 0, size = 9),
    strip.background = element_rect(fill = "grey90", color = "grey50"),
    panel.spacing = unit(0.1, "lines"),
    plot.margin = margin(10, 10, 20, 10)
  ) +
  labs(
    x = "Biomarker", 
    y = "Samples", 
    title = "Heatmap of Biomarkers by Group and Gender"
  )



# Heatmap with means
heatmap_means <- table_clinical_heatmap_imputed %>%
  group_by(group_disease, gender) %>%
  summarise(across(all_of(biomarkers), mean, na.rm = TRUE), .groups = "drop") %>%
  pivot_longer(
    cols = all_of(biomarkers),
    names_to = "biomarker",
    values_to = "value"
  ) %>%
  group_by(biomarker) %>%
  mutate(z_score = scale(value)[,1]) %>%
  ungroup()

heatmap_means <- heatmap_means %>%
  mutate(group_label = paste(group_disease, gender, sep = "\n"))

ggplot(heatmap_means, aes(x = biomarker, y = group_label, fill = z_score)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = round(value, 1)), size = 3, color = "black") +  # Mostrar valores
  scale_fill_gradient2(
    low = "navy", 
    mid = "white", 
    high = "firebrick3", 
    midpoint = 0, 
    name = "Z-score"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 10, face = "bold"),
    axis.text.y = element_text(size = 10, face = "bold"),
    plot.margin = margin(10, 10, 20, 10),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold")
  ) +
  labs(
    x = "\nBiomarker", 
    y = "Group - Gender\n", 
    title = "Average Biomarker Levels by Group and Gender"
  )
```

# Conclusion

Write a concise conclusion (1 paragraphs) summarizing the main differences observed across the three patient groups based on the twelve clinical parameters. Support your statements with evidence from your analyses.

# session info {.unnumbered}

```{r, results='asis',  echo=FALSE, message=FALSE }
sessionInfo()
```
